---
title: "A primer for data analysis"
author: "The R for ABD crew 2020"
date: "29/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# set the root directory somewhere else knitr::opts_knit$set(root.dir = '/tmp')
```

## A simple data analysis example

### Load data

This is a document showcasing a simple R analysis using [Rmarkdown](https://rmarkdown.rstudio.com). We will start by loading a build in dataset called iris

```{r load the file, echo=F}
#this will load the df
data(iris)
head(iris)
```

### Check normality

The dataset contains the following species:  `r unique(iris$Species)`. I will split by species that check for the normality the traits measured in each

```{r split and check normality}
bys<- split(iris, iris$Species)
#make a list to store results
outlist<-list()
for(i in 1:length(bys)){ #i=1
    tmp<-bys[[i]]
    #make a vector to store results
    outnorm<-c()
    for (c in 1:4){ #c=1
      shap<-shapiro.test(tmp[,c])
      outnorm<-c(outnorm, shap$p.value)
    } #for c
    outlist[[i]]<-outnorm
}# for i

tabout<-do.call(rbind, outlist)
colnames(tabout)<-colnames(iris)[1:4]
rownames(tabout)<-names(bys)
tabout
```

### Make plots

It seems that the last phenotype, namely *Petal width*, is not quite normally distributed. We might want to graphically display its distribution (Fig. 1)

```{r plot a weird phenotype, echo=F, fig.cap="**Fig.1** A bunch of histograms made with poor graphical skills"}
par(mfrow=c(3,1))
for(i in 1:length(bys)){
  tmp<-bys[[i]]
  hist(tmp[,4], col="grey")
}#for i

```

Since we are at it, let's make a better graph! Using `ggplot` and `patchwork`to plot the non-normal variable along with a variable with a normal distribution (Fig. 2)

```{r plot of the weird phenotype version 2, echo=T, message=FALSE, warning=FALSE, fig.cap="**Fig.2** A better plot using ggplot"}
#install.packages("ggplot2")
#install.packages("patchwork")
library(ggplot2)
library(patchwork)

g1<-ggplot(iris, aes(y=Petal.Width, x=Species, fill=Species))+geom_boxplot() + labs(tag="A", title="Petal.Width")
g2<-ggplot(iris, aes(x=Petal.Width, fill=Species))+geom_histogram(alpha=0.5, position="identity")

g3<-ggplot(iris, aes(y=Sepal.Width, x=Species, fill=Species))+geom_boxplot() + labs(tag="B", title="Sepal.Width")
g4<-ggplot(iris, aes(x=Sepal.Width, fill=Species))+geom_histogram(alpha=0.5, position="identity")

comb<-(g1 | g2) /  (g3 | g4)
comb

```


### Tranform the data

Since we **really** want to use parametric statistics on *Petal width*, we will go ahead and transform the data (though not always advisable)

```{r transform data, warning=FALSE}
#install.packages("bestNormalize")
library(bestNormalize)
#?bestNormalize

#apply the normalization function to each species
normalized<-lapply(bys, function(x) bestNormalize(x[,4]))
normalized

str(normalized)

```




